---
title: "Chapter 2 Get Started"
output: 
  html_notebook:
    toc: yes
---

```{r}
library(dplyr)
library(socviz)
library(gapminder)
library(tibble)
library(readr)
library(ggplot2)
```


## 2.3 Things to know about R
You do things in R by creating and manipulating named objects.
Every object has a class.
Knowing an object's class tells you a lot about what you can do with it:
```{r}
my_numbers <- c(1,2,3,4)
class(my_numbers)

class(summary)
```
The data set 'Titanic' from the socviz package:
```{r}
glimpse(socviz::titanic) 
class(socviz::titanic)
```
Use 'tibble' as an alternative data frame -- they give more information about the data frame:
```{r}
titanic_tb <- as_tibble(titanic) # as.tibble is deprecated
titanic_tb
```

To see inside an object, ask for its 'structure':
```{r}
str(socviz::titanic)
```
## 2.5 Get data into R
Use 'read_csv()' to read in comma-separated data.
'read_csv()' will not classify variables as factors unless you tell it to.
```{r}
?read_csv
```
## 2.6 Make your first figure
```{r}
glimpse(gapminder)
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point()
```

