---
title: "Chapter 3 Make a Plot"
output: 
  html_notebook:
    toc: yes
---

```{r}
# knitr::opts_chunk$set(fig.width = 12, fig.height = 9)
```



```{r}
library(dplyr)
library(socviz)
library(gapminder)
library(tibble)
library(readr)
library(ggplot2)
library(scales)
```

## 3.1 How Ggplot Works
## 3.2 Tidy Data
On page 57 there is a good comparison of a 'wide' table and that table converted into a 'long' table.
The 'wide' table where some variables are spread across the columns of the table (e.g years across the columns)
## 3.3 Mappings link data to things you see
The mapping argument of ggplot (ggplot(aes( )...)) aes() is a function. aes() links variables to things.
## 3.4 Build your plots layer by layer
Adding a point layer and a smooth layer (note using mgcv::gam which is the default for geom_smooth()):
```{r}
?mgcv::`mgcv-package`
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point() +
  geom_smooth(method = mgcv::gam)
```

'gam' means that it has fit a generalized additive model. geom_smooth is a function with other arguments. Try
setting its method argument to 'lm' for linear model (note: 'lm' is a function as well):
```{r}
?geom_smooth
?lm
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point() +
  geom_smooth(method = lm)
```
Data bunched up on left side. x scale would look better if we used a log scale and the mgcv::gam smoothing method.
Note the points are repositioned to a more linear trend shape because of the rescaling.
```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point() +
  geom_smooth(method = mgcv::gam) +
  scale_x_log10()
```
Decorating the plot: put dollar signs on x axis using 'scales' package. The 'labels' argument modifies the axis text.
```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point() +
  geom_smooth(method = mgcv::gam) +
  scale_x_log10(labels = scales::dollar)
```
## 3.5 Mapping aesthetics vs setting them
The aes property 'color' will represent the variable 'continent':
```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point()
```
aes is for linking a variable to 'color', 'size', etc.  To actually change point colors, define it in a geom_.
See the below help for list of aesthetics that geom_point understands
```{r}
?geom_point
```
Additional decorations with 'alpha' and 'labs()':
```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = mgcv::gam) +
  scale_x_log10(labels = scales::dollar) +
  labs(
    x = "GDP Per Capita",
    y = "Life Expectency in Years",
    title = "Economic Growth and Life Expectancy",
    subtitle = "Data points are country-years",
    caption = "Source: Gapminder."
  )
```
All geom's inherit from the original aes settings (note: using 'fill' for the error bars):
```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent, fill = continent)) +
  geom_point() +
  geom_smooth(method = stats::loess) +
  scale_x_log10()
```

## 3.6 Aesthetics can be mapped per geom_
By default geom_'s inherit their mappings from the ggplot() function.  We can change this by specifying different aethetics for each geom_.  We use same aes() expression in the ggplot() call but now use it in the geom_ function as well, specifying the aes() we want for each one. 
Example--set the aes() of geom_point to 'aes(color = log(pop)):
Note the use of 'theme() function for controlling the legend position.
```{r, fig.width = 12, fig.height = 9}
?ggplot2::theme
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(aes(color = log(pop))) +
  scale_x_log10() + 
  theme(legend.position = "top")
```
## 3.7 Save your work
Changing the size of plots in .Rmd documents: Set the following option in the first code chunck:
'knitr::opts_chunk$set(fig.width gdpPercap, lifeExp= 8, fig.height = 5)' or individual chunck as above.
To save the most recently displayed figure (available: *.png, *.pdf, *.jpg, *.svg):
```{r}
ggsave(filename = "figures/gdpPercap_lifeExp_log.jpg", height = 4, width = 6, units = "in")
ggsave(filename = "figures/gdpPercap_lifeExp_log.svg")
```
## 3.8 Where to go next





