---
title: "Chapter 5 Graph Tables, Add Labels, Make Notes"
output: 
  html_notebook:
    toc: yes
---



```{r}
library(dplyr)
library(tibble)
library(ggplot2)
library(scales)
library(socviz)
library(forcats)
library(ggrepel)
```

Using the package 'dplyr' to get data into shape before using ggplot.
Expand the number of geom_'s and how to choose among them.
Learn more about scale, guide, and theme functions

Doing the same thing:
  1. Define a set of tidy data.
  2. Mapping of variables to aesthetic elements.
  3. Select a particular type of graph (geom_)

## 5.1 Use pipes to summarize data
Using dplyr package and socviz::gss_sm dataset build a summary table with percentages of religious preferences grouped within region.

The main functions to use from dplyr package: group_by(), filter(), select(), mutate(), summarize().
```{r}
glimpse(socviz::gss_sm)
```
Note: there are some 'NA' or missing values in the 'religion' factor:
```{r}
sum(is.na(gss_sm$religion))
```
So using forcats::fct_explicit_na() function, give missing values an explicit factor level, ensuring that they appear in summaries and on plots.
```{r}
gss_sm_missing <- gss_sm %>% 
  mutate(
    religion = forcats::fct_explicit_na(gss_sm$religion, na_level = "(Missing)")
  )
# check for NA's under religion again
sum(is.na(gss_sm_missing$religion))
levels(gss_sm_missing$religion)
```


```{r}
(rel_by_region <- gss_sm_missing %>% 
  group_by(bigregion, religion) %>% 
  summarise(N = n()) %>% 
  mutate(
    frac = N/sum(N),
    pct = round(frac * 100, 0)
  )
)
```
Note that 'summarise()' retains the 'bigregion' and 'religion' variables; adds a new variable 'N' and drops all the other variables.  'mutate()' adds two new variables -- 'frac' and 'pct'.

Check that the percentages add up to 100% for each 'bigregion':
```{r}
(rel_by_region %>% 
  group_by(bigregion) %>% 
  summarize(
    total_pct = sum(pct)
  )
)
```
Since we are working directly with percentages, we can use 'geom_col()' instead of 'geom_bar()':
```{r}
ggplot(rel_by_region, aes(bigregion, pct, fill = religion)) +
  geom_col(position = "dodge2") +
  labs(
    x = "Regioin",
    y = "Percent",
    fill = "Religion"
  ) +
  theme(legend.position = "top")
```
Note: 'dodge2' adds padding = 0.1, whereas 'dodge' does not add padding among the bars.
Can improve chart -- above chart has too many bars -- hard to compare bars.  Use faceting to remove the 'fill = religion' legend and make the chart easier to read.  With facetiing for 'bigregion', put 'religion' on the x axis and 'pct' on the y axis and then flip the axis':
```{r, fig.width=9}
ggplot(rel_by_region, aes(religion, pct, fill = religion)) +
  geom_col(position = "dodge2") +
  labs(x = NULL, y = "Percent") +
  guides(fill = FALSE) +
  coord_flip() +
  facet_grid(~bigregion)
```
## 5.2 Continuous variables by group or category
New dataset: 'socviz::organdata'
```{r}
glimpse(socviz::organdata)
```
Plot 'year' by 'donors':
```{r, fig.width=12}
ggplot(organdata, aes(year, donors)) +
  geom_line(aes(group = country)) +
  facet_wrap(~country)
```
Looking at country variation without looking at the yearly trend using geom_boxplot():
```{r, fig.width=12}
ggplot(organdata, aes(country, donors)) +
  geom_boxplot() +
  coord_flip()
```
Reorder donations from high to low using reorder() function.  reorder() takes two arguments: the varible to be reordered and the variable to base the reordering on:
```{r}
ggplot(organdata, aes(x = reorder(country, donors, na.rm = TRUE), y = donors)) +
  geom_boxplot() +
  labs(x = NULL) +
  coord_flip()
```
geom_boxplot can accept a fill variable:
```{r}
ggplot(organdata, aes(x = reorder(country, donors, na.rm = TRUE), y = donors, fill = world)) +
  geom_boxplot() +
  labs(x = NULL) +
  coord_flip() +
  theme(legend.position = "top")
```
With the number of observations in each country small, we can geom_point() (or geom_jitter() for overlapped points) instead of using geom_boxplot():
```{r}
ggplot(organdata, aes(x = reorder(country, donors, na.rm = TRUE), y = donors, color = world)) +
  geom_jitter(position = position_jitter(width = 0.15)) +
  labs(x = NULL) +
  coord_flip() +
  theme(legend.position = "top")
```
Grouping organdata by consent_law and country variables and summarizing with a number of statistics (using dplyr):
```{r}
(by_country <- organdata %>% 
  group_by(consent_law, country) %>% 
  summarize(
    donors_mean = mean(donors, na.rm = TRUE),
    donors_sd = sd(donors, na.rm = TRUE),
    gdp_mean = mean(gdp, na.rm = TRUE),
    health_mean = mean(health, na.rm = TRUE),
    roads_mean = mean(roads, na.rm = TRUE),
    cerebvas_mean = mean(cerebvas, na.rm = TRUE)
      
  )
)
```
An alternative where we summarize all the numeric variables:
```{r}
(by_country <- organdata %>% 
  group_by(consent_law, country) %>% 
  summarize_if(is.numeric, list(mean = mean, sd = sd), na.rm = TRUE) %>% 
  ungroup() # so that the result is a plain table
)
```
Plot donors_mean by country with Donor and color of consent_law:
```{r}
ggplot(by_country, aes(donors_mean, y = reorder(country, donors_mean), color = consent_law)) +
  geom_point(size = 3) +
  labs(
    x = "Donor Procurement Rate",
    y = "",
    color = "Consent Law"
  ) +
  theme(legend.position = "top")
```
Cleveland dotplots are generally preferred to bar or column charts.  
Faceting by consent_law to remove the legend and color and scales = 'free_y':
```{r, fig.height=8, fig.width=9}
ggplot(by_country, aes(donors_mean, y = reorder(country, donors_mean))) +
  geom_point(size = 3) +
  facet_wrap(~ consent_law, scales = "free_y", ncol = 1) +
   labs(
    x = "Donor Procurement Rate",
    y = ""
  )
```
Use geom_pointrange to plot sd around the mean:
```{r}
ggplot(by_country, aes(x = reorder(country, donors_mean), y = donors_mean)) +
  geom_pointrange(aes(ymin = donors_mean - donors_sd, ymax = donors_mean + donors_sd)) +
  labs(x = "", y = "Donor Procuremen Rate") +
  coord_flip()
```
## 5.3 Plot text directly
Using geom_text().
```{r}
ggplot(by_country, aes(roads_mean, donors_mean)) +
  geom_point() +
  geom_text(aes(label = country))
```
Adjust text position (hjust = 0 is left justify; hjust = 1 is right justify):
```{r}
ggplot(by_country, aes(roads_mean, donors_mean)) +
  geom_point() +
  geom_text(aes(label = country), hjust = 0)
```
Use package ggrepel for more flexible text labelling. Using a new dataset -- socviz::elections_historic:
```{r}
socviz::elections_historic %>% select(2:7)
```
```{r, fig.width=10, fig.height=10}
p.title <- "Presidential Elections: Popular and Electoral College Margins"
p.subtitle <- "1824 - 2016"
p.caption <- "Data for 2016 are provisional"
x_label <- "Winner's Share of Popular Vote"
y_label <- "Winner's Share of Electoral College Vote"
ggplot(elections_historic, aes(popular_pct, ec_pct, label = winner_label)) +
  geom_hline(yintercept = 0.5, size = 1.4, color = "gray80") +
  geom_vline(xintercept = 0.5, size = 1.4, color = "gray80") +
  geom_text_repel() +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = x_label, y = y_label, title = p.title, subtitle = p.subtitle, caption = p.caption)
```
## 5.4 Label outliers
Selecting points of interest.
```{r}
ggplot(by_country, aes(gdp_mean, health_mean)) +
  geom_point() +
  geom_text_repel(data = subset(by_country, gdp_mean > 25000 | health_mean < 1500 | country %in% "Belgium"), aes(label = country))
```
Creating a dummy variable:
```{r}
organdata$ind <- organdata$ccode %in% c("Ita", "Spa") & organdata$year > 1998
ggplot(organdata, aes(x = roads, y = donors, color = ind)) +
  geom_point() +
  geom_text_repel(data = subset(organdata, ind), aes(label = ccode)) +
  guides(label = FALSE, color = FALSE)

```
## 5.5 Write and draw in the plot area
Using annotate() function.
```{r}
ggplot(organdata, aes(roads, donors)) +
  geom_point() +
  annotate(geom = "text", x = 91, y = 33, label = "A surprisingly high \n recovery rate", hjust = 0)
```
## 5.6 Understanding scales, guides, themes
Set the aestheic to 'shape'
```{r, fig.width=8,fig.height=8}
ggplot(organdata, aes(roads, donors, shape = world)) +
  geom_point(size = 3)
```
Controlling the tick marks:
```{r, fig.width=8,fig.height=8}
ggplot(organdata, aes(roads, donors, shape = world)) +
  geom_point(size = 3) +
  scale_x_log10() +
  scale_y_continuous(breaks = c(5, 15, 25), labels = c("Five", "Fifteen", "Twenty-Five"))
```





