---
title: "Chapter 4 Show the Right Numbers"
output: 
  html_notebook:
    toc: yes
---



```{r}
library(dplyr)
library(magrittr)
library(socviz)
library(gapminder)
library(tibble)
library(readr)
library(ggplot2)
library(scales)
```
## 4.2 Grouped data and the "Group" aesthetic
Plot the gapminder data grouped by country across the years with a line chart:
```{r}
ggplot(gapminder, aes(year, gdpPercap)) +
  geom_line(aes(group = country))
```
## 4.3 Facet to make small multiples
```{r}
ggplot(gapminder, aes(year, gdpPercap)) +
  geom_line(aes(group = country)) +
  facet_wrap(~continent)
```
Enhanced facet:
```{r, fig.width=12}
ggplot(gapminder, aes(year, gdpPercap)) +
  geom_line(color = "gray70", aes(group = country)) +
  geom_smooth(size = 1.1, method = stats::loess, se = FALSE) +
  scale_y_log10(labels = scales::dollar) +
  facet_wrap(~continent, ncol = 5) +
  labs(
    x = "Year",
    title = "GDP per capita on Five Countries" 
  )
```
Using the dataset 'gss_sm'
```{r}
glimpse(gss_sm)
```
Faceting gss_sm by sex and race
```{r}
ggplot(gss_sm, aes(age, childs)) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  facet_grid(sex ~ race)
```
## 4.4 Geoms can transform data
Specify 'group = 1' when we want to total number of observations to compute a proportion (i.e. ..prop..)
```{r}
ggplot(gss_sm, aes(bigregion)) +
  geom_bar(aes(y = ..prop.., group = 1))
```
What is the religious preference?
```{r}
table(gss_sm$religion)
```
Mapping two aesthetics to the same variable without the ledgend:

```{r}
ggplot(gss_sm, aes(x = religion, fill = religion)) +
  geom_bar() +
  guides(fill = FALSE)
```
## 4.5 Frequency plots the slightly awkward way
The distributions of religions across different regions of US - the 'religion' variable broken down proportionally within 'bigregion' (the default plot is a stacked bar chart):
```{r}
ggplot(gss_sm, aes(x = bigregion, fill = religion)) +
  geom_bar()
```
Using 'position' = "dodge" the y axis reverts back to count:
```{r}
ggplot(gss_sm, aes(x = bigregion, fill = religion)) +
  geom_bar(position = "dodge")
```
Using '..prop..':
```{r}
ggplot(gss_sm, aes(x = bigregion, fill = religion)) +
  geom_bar(position = "dodge", aes(y = ..prop.., group = religion))
```
```{r}
ggplot(gss_sm, aes(x = religion)) +
  geom_bar(position = "dodge", aes(y = ..prop.., group = bigregion)) +
  facet_wrap(~bigregion, ncol = 2)
```
Using the 'dplyr' first before plotting will make the above attempts easier.

## 4.6 Histograms and density plots
A histogram chops a continuous variable into bins. geom_histogram will compute the bins for us.  Looking at a histogram of continuous variable 'area' from dataset 'midwest':
```{r}
ggplot(midwest, aes(x = area)) +
  geom_histogram()
```
```{r}
ggplot(midwest, aes(x = area)) +
  geom_histogram(bins = 18)
```
Calculate a kernal density estimate of the underlying distribution:
```{r}
ggplot(midwest, aes(x = area)) +
  geom_density()
```
## 4.7 Avoid transformations when necessary
You all ready have the proportions or counts data and don't need ggplot to perform statistics. To turn off ggplot doing stats then enter: stat = "identity" in the geom_bar.
```{r}
ggplot(titanic, aes(x = fate, y = percent, fill = sex)) +
  geom_bar(position = "dodge", stat = "identity") +
  theme(legend.position = "top")

```








