---
title: "Chapter 7 Draw Maps"
output: 
  html_notebook:
    toc: yes
---

```{r}
library(dplyr)
library(tibble)
library(ggplot2)
library(scales)
library(socviz)
library(maps)
library(ggthemes)
```

## 7.1 Map U.S. state-level data
```{r}
?socviz::election
glimpse(socviz::election)
```
Plot 'r_points' vs 'state' by 'census' (faceted):
```{r, fig.width=8, fig.height=12}
?ggplot2::facet_wrap
party_colors <- c("#2E74C0", "#CB454A")
socviz::election %>% 
  filter(!election$st %in% c("DC")) %>% 
  ggplot(aes(r_points, reorder(state, r_points), color = party)) +
  geom_vline(xintercept = 0, color = "gray30") +
  geom_point(size = 2) +
  scale_color_manual(values = party_colors) +
  scale_x_continuous(
    breaks = c(-30, -20, -10, 0, 10, 20, 30, 40),
    labels = c("30\n(Clinton)", "20", "10", "0", "10", "20", "30", "40\n(Trump)")
  ) +
  facet_wrap(~ census, ncol = 1, scales = "free_y") +
  guides(color = FALSE) +
  labs(x = "Point Margin", y = NULL) +
  theme(axis.text = element_text(size = 8))
```
Set up a dataframe with drawing info for states:
```{r}
us_states <- map_data("state")
glimpse(us_states) 
```
Plot the states:
```{r}
ggplot(us_states, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black")
```
Color the states by setting 'fill' to 'region':
```{r}
ggplot(us_states, aes(long, lat, group = group, fill = region)) +
  geom_polygon(color = "gray90", size = 0.1) +
  guides(fill = FALSE)
```
Using the Albers projection for a more accurate US map:
```{r}
ggplot(us_states, aes(long, lat, group = group, fill = region)) +
  geom_polygon(color = "gray90", size = 0.1) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  guides(fill = FALSE)
```
left-join the election data with the map data:
```{r}
election$region <- tolower(election$state)
us_states_elec <- left_join(us_states, election)
glimpse(us_states_elec)
```
With the election data merged with map data, plot the states with fill = 'party':
```{r}
ggplot(us_states_elec, aes(x = long, y = lat, group = group, fill = party)) +
  geom_polygon(color = "gray90", size = 0.1) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  guides(fill = FALSE)
```
Remove the grids, use party colors, add a legend:
```{r}
party_colors <- c("#2E74C0", "#CB454A")
ggplot(us_states_elec, aes(x = long, y = lat, group = group, fill = party)) +
  geom_polygon(color = "gray90", size = 0.1) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  scale_fill_manual(values = party_colors) +
  labs(title = "Election Results 2016", fill = NULL) +
  ggthemes::theme_map()
```
Plot a continuous variable -- fill = 'pct_trump':
```{r}
ggplot(us_states_elec, aes(x = long, y = lat, group = group, fill = pct_trump)) +
  geom_polygon(color = "gray90", size = 0.1) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  scale_fill_gradient(low = "white", high = "#CB454A") +
  labs(title = "Trump vote", fill = "Percent") +
  ggthemes::theme_map()
```

