---
title: "Fun With the R Grid Package"
author: "Rick Dean"
date: "09/06/2020"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    code_folding: hide
    css: ../../../style.css
---

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(grid)
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 10)
```

The following R scripts and notes are sourced from [Fun With the R Grid Package](http://jse.amstat.org/v18n3/zhou.pdf), an article by Lutong Zhou and W. John Braun

# 1 Introduction
"We have found that the grid package is an interesting and surprisingly simple way to do a
lot of things that are either difficult or impossible using the more traditional base graphics."

"The grid package is an R package, developed by Paul Murrell in the 1990s."

"We will describe the very basic ideas of the grid package, introducing viewports, editing graphic objects
(grobs), and demonstrating how grid can be an alternative to traditional graphics. The
second purpose of the paper is to demonstrate that, in addition to what is already available
in the *lattice* package, the grid package offers flexibility which can be exploited in a variety
of ways to visualize data."

## 1.1 Using Viewports
<div class="task">Task: Create a viewport centered at (.5,.5), with width .9 and height .9 relative to the graphics window.</div>

```{r}
main_vp <- grid::viewport(x=0.5,y=0.5,width = 0.9,height = 0.9,name = "main_vp")
grid::grid.show.viewport(v = main_vp)
```
A viewport name can be any string, but *grid* uses the reserved name "ROOT" for the top-level viewport.

<div class="task">Task: Create a new page, tell R to use the viewport by calling `grid::pushViewport()`, add rect and circle. </div>
```{r}
grid::grid.newpage(recording = FALSE)
grid::pushViewport(main_vp)
grid::grid.rect(gp=gpar(lty="dashed",col="red"))
grid::grid.circle(x=0.6,y=0.4,r=0.3)
```
By default, the coordinates of the lower left corner of the viewport are (0,0), and the upper
right corner has coordinates (1,1).

<div class="task">Task: Create a second(smaller) viewport "small_vp" to be placed within the "main_vp" viewport. </div>

```{r}
small_vp <- viewport(x=0.5,y=0.75,width = 0.6,height = 0.3,name = "small_vp")
```

<div class="task">Task: Create a new page with "main_vp", "small_vp" and place a circle and retangle in the "small_vp".</div>

```{r}
grid::grid.newpage(recording = FALSE)
grid::pushViewport(main_vp)
grid::grid.rect(gp=gpar(lty="dashed",col="red"))
grid::pushViewport(small_vp)
grid::grid.rect(gp=gpar(lty="dashed",col="blue"))
grid::grid.circle(gp=gpar(col="purple"))
```
The specifications of "small_vp" `(x=0.5,y=0.75,width = 0.6,height = 0.3)` are relative to the viewport that it has been pushed into (i.e. "main_vp").

We can change our focus from one viewport to another and back, using the *pushViewport()* and *upViewport()* functions. The commands *downViewport()* and *popViewport()* also change our focus.  After pushing "small_vp" it is now the current viewport inside "main_vp". Also, when specifying a viewport name in *downviewport()* or *seekViewport()*, it is possible to provide a viewport path, which consists of several names concatenated using the separator(currently ::).

<div class="task">Task: Change viewport focus from "small_vp" to "main_vp" and add a circle.</div>  
```{r}
grid::grid.newpage(recording = FALSE)
grid::pushViewport(main_vp)
grid::grid.rect(gp=gpar(lty="dashed",col="red"))
grid::pushViewport(small_vp)
grid::grid.rect(gp=gpar(lty="dashed",col="blue"))
grid::grid.circle(gp=gpar(col="purple"))

grid::upViewport(n=1)

# add circle centered at (.5,.5) and radius of 1
grid::grid.circle(gp=gpar(col="purple", alpha = 0.5))
```
`grid::upViewport(n=1)`moves up 1 generation.

<div class="task">Task: Nest 30 viewports and place a "stickperson" within the 5th, 20th, and 30th viewports.</div>  
1. Define the "stickperson" function:
```{r}
stickperson <- function(){
  grid.circle(x=.5, y=.8, r=.1, gp=gpar(fill="yellow"))
  grid.lines(c(.5,.5), c(.7,.2)) # vertical line for body
  grid.lines(c(.5,.7), c(.6,.7)) # right arm
  grid.lines(c(.5,.3), c(.6,.7)) # left arm
  grid.lines(c(.5,.65), c(.2,0)) # right leg
  grid.lines(c(.5,.35), c(.2,0)) # left leg
}
```
2. Create a new page with 30 nested viewports and "stickperson"'s:
```{r}
grid::grid.newpage(recording = FALSE)
pushViewport(viewport())
grid::grid.lines(c(0.05,0.95), c(0.95,0.05))
grid::grid.lines(c(0.05,0.95), c(0.05,0.95))
for(i in 1:30){
  vp <- viewport(height = 0.9, width = 0.9)
  pushViewport(vp)
  grid.rect(gp=gpar(alpha = 0.1))
  # stickperson 1:
  if(i == 5){
    pushViewport(viewport(x = 0.8))
    stickperson()
    upViewport()
  }
  # stickperson 2:
  if(i == 20){
    pushViewport(viewport(x = 0.2))
    stickperson()
    upViewport()
  }
  # stickperson 3:
  if(i == 30){
    stickperson()
  }
}
```

## 1.2 Using Viewports to Display Data
The data set:
```{r}
escape_prop <- c(0.24, 0.28, 0.28, 0.33, 0.33, 0.32, 0.30, 0.21, 0.30, 0.28, 
                 0.17, 0.27, 0.21, 0.18, 0.22, 0.21, 0.19, 0.17, 0.17, 0.15,
                 0.25, 0.19, 0.19, 0.22, 0.21, 0.18, 0.24, 0.23, 0.27, 0.16,
                 0.17, 0.22, 0.17, 0.25, 0.19, 0.25, 0.12, 0.17, 0.22, 0.22)
nfires <- c(953, 620, 584, 839, 1415, 1180, 656, 408, 872, 965, 
            853, 1492, 951, 772, 1541, 1114, 479, 860, 1166, 1208, 
            657, 1140,1223, 1275, 489, 932, 1096, 1378, 1033, 889, 
            1046, 818, 1213,782, 962, 1666, 2017, 1689, 1885, 1435)
nfires_scaled <- nfires/max(nfires)
index <- (1:40)/41
```
<div class="task">Task: Plot the fire data using *grid*.</div>  
```{r}
grid::grid.newpage()
grid::pushViewport(viewport(width = .9, height = .9))
grid::pushViewport(viewport(y=0.75, width = .9, height = .9))
for(i in 1:40){
  vp <- viewport(x=index[i], y = escape_prop[i], height = 0.03, width = 0.03)
  grid::pushViewport(vp)
  grid::grid.circle(r = sqrt(nfires_scaled[i]/pi))
  grid::upViewport()
}
grid::grid.xaxis(at = c(0, index[c(10,20,30,40)]), label = seq(1960,2000,10))
grid::grid.yaxis(at=seq(0,0.5,0.1))
grid::grid.text("Proportion of Escaped Fires", y = 0.6)
```

## 1.3 Graphics Objects and Editing
An important feature of grid graphics is the graphical object (also called a *grob*). There are two ways to interact with a grob:
  1. Drawing it immediately -- `grid::grid.rect()` creates a grob and draws it
  2. Create it but don't draw -- `grid::rectGrob()` creates a grob but does not draw it; to draw it use `grid::grid.draw()`
  
Modify a grob by using `grid::grid.edit()` and `grid::editGrob()`

<div class="task">Task: Create and edit some rectangle grobs.</div>  
  1. Create but don't draw a rectangle and name it "gr_rect_1"
```{r}
gr_rect_1 <- grid::rectGrob(width = 0.1, height = 0.1, name = "gr_rect_1") #note: x = 0.5, y = 0.5 by default
```

  2. Using `grid::editGrob()` copy "gr_rect_1" centered at (0.2, 0.6) and draw it
```{r}
grid::grid.newpage()
gr_copy_1 <- grid::editGrob(grob = gr_rect_1, vp = viewport(x = 0.2, y = 0.6), name = "gr_copy_1")
grid::grid.draw(gr_copy_1)
```
  3. Create and draw a second copy centered at (0.7, 0.75:
```{r}
grid::grid.newpage()
gr_copy_2 <- grid::editGrob(grob = gr_rect_1, vp = viewport(x = 0.7, y = 0.75), name = "gr_copy_2")
grid::grid.draw(gr_copy_1)
grid::grid.draw(gr_copy_2)
```
 3. Create and draw a third copy centered at (0.5, 0.4: 
```{r}
grid::grid.newpage()
gr_copy_3 <- grid::editGrob(grob = gr_rect_1, vp = viewport(x = 0.5, y = 0.4), name = "gr_copy_3")
grid::grid.draw(gr_copy_1)
grid::grid.draw(gr_copy_2)
grid::grid.draw(gr_copy_3)
```
<div class="task">Task: Use `grid::grid.edit()` to rotate the rectangles via using `grid::viewport()`.</div>  
```{r}
grid::grid.newpage()
gr_copy_1 <- grid::editGrob(grob = gr_copy_1, vp = viewport(x = 0.2, y = 0.6, angle = 30))
gr_copy_2 <- grid::editGrob(grob = gr_copy_2, vp = viewport(x = 0.7, y = 0.75, angle = 63))
gr_copy_3 <- grid::editGrob(grob = gr_copy_3, vp = viewport(x = 0.5, y = 0.4, angle = 72))
grid::grid.draw(gr_copy_1)
grid::grid.draw(gr_copy_2)
grid::grid.draw(gr_copy_3)
```

<div class="task">Task: Create a function `draw_star()` for drawing a single star.</div>
```{r}
draw_star <- function(x = 0.5, y = 0.5, width = 0.25, height = 0.25, angle = 0, name = "a_star", fill = "purple"){
  grid::pushViewport(vp = viewport(x = x, y = y, width = width, height = height, angle = angle, name = name))
  b2 <- sqrt(1/cos(36 * pi/180)^2-1)/2
  b3 <- sin(72*pi/180)/(2*(1+cos(72*pi/180))) - (1-sin(72*pi/180))/2
  a_triangle <- grid::polygonGrob(c(0, 0.5,1), c(b3, b2+b3, b3), 
                                  gp = gpar(fill = fill, 
                                  col = 0)
                                 )
  for(i in 0:2){
    grid::pushViewport(vp = viewport(angle = 72 * i, name = paste0("triangle_",i)))
    grid::grid.draw(a_triangle)
    grid::upViewport()
  }
  grid::upViewport()
}
grid.newpage()
draw_star(name = "star_1", fill = "yellow")
grid.ls(viewports = TRUE)
```

<div class="task">Task: Create rotated and scaled stars on a blue background.</div>  
```{r}
grid::grid.newpage()
# create the background viewport
grid::grid.rect(gp = gpar(fill = "blue"))

#draw star 1
draw_star(x=0.2, y=0.2, width = 0.25, height = 0.25, angle = 40, fill = "yellow", name = "star_1")

#draw star 2
draw_star(x=0.8, y=0.8, width = 0.3, height = 0.3, angle = 90, fill = "red", name = "star_2")

#draw star 3
draw_star(x=0.7, y=0.3, width = 0.2, height = 0.2, angle = 130, fill = "green", name = "star_3")

#draw star 4
draw_star(x=0.3, y=0.7, width = 0.15, height = 0.15, angle = 210, fill = "purple", name = "star_4")
```

## 1.4 gTree and gList
The function `grid::gList()` allows us to create a grob that contains a list of grobs.
<div class="task">Task: Redefine the above function `draw_star()` to `create_star()` that returns a single grob containing a list of triangle grobs.</div>

```{r}
create_star <- function(x = 0.5, y = 0.5, width = 0.25, height = 0.25, angle = 0, name = "a_star", fill = "purple") {
  b2 <- sqrt(1/cos(36 * pi/180)^2-1)/2
  b3 <- sin(72*pi/180)/(2*(1+cos(72*pi/180))) - (1-sin(72*pi/180))/2
 
  starGrobs <- grid::gList(
    grid::polygonGrob(
      x = c(0, 0.5,1), 
      y = c(b3, b2+b3, b3), 
      gp = grid::gpar(fill = fill, col = 0),
      vp = grid::viewport(angle = 0, name = paste0("triangle_",0))
    ),
    grid::polygonGrob(
      x = c(0, 0.5,1), 
      y = c(b3, b2+b3, b3), 
      gp = grid::gpar(fill = fill, col = 0),
      vp = grid::viewport(angle = 72, name = paste0("triangle_",1))
    ),
    grid::polygonGrob(
      x = c(0, 0.5,1), 
      y = c(b3, b2+b3, b3), 
      gp = grid::gpar(fill = fill, col = 0),
      vp = grid::viewport(angle = 144, name = paste0("triangle_",2))
    )
  )
  starTree <- grid::gTree(children = starGrobs, name = name, vp = grid::viewport(x = x, y = y, width = width, height = height, angle = angle))
 
  return(starTree)
}

starTree <- create_star(x = .5, y = .5, width = 0.25, height = 0.25, angle = 30)
grid::grid.newpage()
grid::grid.draw(starTree)
```

## 1.5 Other Scales

*native* - locations and sizes are relative to the x and y scales for the current viewport
*npc* (the default) - Normalized Parent Coordinates. Treats the bottom left corner of the current viewport as the location of (0,0) and top right as (1,1).

Other scales: cm, inches, mm, points, lines

